---
title: "zx_final_project"
author: "Zihan Xiong"
date: "2026-05-10"
output: html_document
---
**Note that this rmd code file only included my entire analysis and a few sentences for self reminded of what I should say in the final report. Please see the final report for the full results and discussion.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE}
library(dplyr)
library(caret)
library(glmnet)
library(ggplot2)
library(tidymodels)
library(patchwork)
library(gtsummary)
library(corrplot)
```


# Load data and Preparation
```{r}
load("severity.RData")
set.seed(2527)
my_dat <- dat[sample(nrow(dat),1000), ]
sum(is.na(dat))
```

There are no NAs in this dataset.

## Data preparation and wrangling
```{r}
my_dat$id <- NULL

my_dat <- my_dat %>%
  mutate(
  gender = factor(gender,
                  levels = c(0,1),
                  labels = c("Female","Male")),
                  
  race = factor(race,
                  levels = c(1,2,3,4),
                  labels = c("White","Asian","Black","Hispanic")),
                
  smoking = factor(smoking,
                   levels = c(0,1,2),
                   labels = c("NeverSmoked","FormerSmoker","CurrentSmoker")),
                   
  diabetes = factor(diabetes,
                   levels = c(0,1),
                   labels = c("No","Yes")),

  hypertension = factor(hypertension,
                   levels = c(0,1),
                   labels = c("No","Yes")),
  
  vaccine = factor(vaccine,
                   levels = c(0,1),
                   labels = c("NotVaccinated","Vaccinated")),
  
  severity = factor(severity,
                    levels = c(0,1),
                    labels = c("NotSevere","Severe"))
  )
str(my_dat)

```

```{r}
table(my_dat$severity) # want to check the proportion of severe vs not severe
```

Not very severe imbalance, use AUC and stratified split.

### Table1. Summary Statistics
Vaccine, hypertension, and SBP might be crutial predictors. Beware that there might be collinearity between height, weight and BMI. Might keep BMI in the future.
```{r}
my_dat |>
  tbl_summary(
    by = severity,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2
  ) |>
  add_p() |>
  add_overall() |>
  modify_header(label ~ "Variable") |>
  bold_labels()
```
## Figure1.feature plot
```{r}
numeric_vars <- c("age","weight","height","bmi","SBP","LDL","depression")

featurePlot(
  x=my_dat[, numeric_vars],
  y=my_dat$severity,
  scales=list(x = list(relation = "free"),
                  y = list(relation = "free")),
    plot = "density",
    pch  = "|",                           
    auto.key = list(columns = 2)   
)
```

## Figure2 vaccine and hypertension 
```{r}
p_vaccine <- ggplot(my_dat,aes(x=vaccine,fill = severity)) +
  geom_bar(position = "fill") + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(x = "Vaccination status", y = "Proportion", fill = "Severity") +
    theme_minimal()

  p_hyper <- ggplot(my_dat, aes(x = hypertension, fill = severity)) +
    geom_bar(position = "fill") +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(values = c("steelblue", "darkorange")) +
    labs(x = "Hypertension", y = NULL, fill = "Severity") +
    theme_minimal()
  
p_vaccine + p_hyper + plot_layout(guides = "collect")
```

## Figure3 matrix
```{r}
numeric_vars <- c("age","weight","height","bmi","SBP","LDL","depression")

cor_mat <- cor (my_dat[, numeric_vars])

corrplot(cor_mat,
         method = "color",
         type = "upper",        
         addCoef.col = "black",   
         tl.col = "black",      
         tl.srt = 45,             
         number.cex = 0.8)
```
keep height and weight, and exclude bmi -> moderate collinearity.

# data split and cross validation

## split
```{r}
set.seed(2527)

trainRows <- createDataPartition(my_dat$severity, p=0.8, list = FALSE)

train_dat <- my_dat[trainRows, ]
test_dat <- my_dat[-trainRows, ]

prop.table(table(train_dat$severity))
prop.table(table(test_dat$severity))
```
## CV control
```{r}
ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)
```

10-fold cross validation

# Model training

## Simpler Models

Exclude height and weight first.
```{r}
train_dat <- train_dat |> select(-height, -weight)
test_dat  <- test_dat  |> select(-height, -weight)
```

1. Logistic regression (as the baseline?)
```{r}
set.seed(2527)

model1 <- train(severity ~.,
                data = train_dat,
                method = "glm",
                metric = "ROC",
                trControl = ctrl
)

model1
```
2. penalized logistics
```{r}
model2 <- train(severity ~.,
                data = train_dat,
                method = "glmnet",
                metric = "ROC",
                trControl = ctrl,
                tuneGrid = expand.grid(
                  alpha = seq(0,1,length=6),
                  lambda = exp(seq(-8,1,length = 50))
                )
)
model2
plot(model2)
coef(model2$finalModel, model2$bestTune$lambda)
```

Hint?: no need to exclude more predictors for very small alpha and beta. AUC is very close to model1's. (AUC=0.89)

3. GAM